---
title: "Untitled"
author: "Michael Rose"
date: "January 10, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(stringr)
library(microbenchmark)
```

### String Basics

```{r}

string1 <- "This is a string"
string2 <- 'To put a "quote" inside a string, use single quotes'

double_quote <- "\"" # or '"'
double_quote

single_quote <- '\''

literal_backslash <- "\\"

writeLines(double_quote)

x <- c("\"", "\\")
writeLines(x)

x <- "\u00b5"
writeLines(x)

c("one", "two", "three")
```

### String Length

```{r}
str_length(c("a", "R for Data Science", NA))
```

### Combining Strings

```{r}
str_c("x", "y")
str_c("x", "y", "z")

# to control how they are separated

str_c("x", "y", sep = ", ")

# to print as NA

x <- c("abc", NA)

# without NA in x

str_c("|-", x, "-|")

# with NA in x

str_c("|-", str_replace_na(x), "-|")

str_c("prefix-", c("a", "b", "c"), "-suffix")

# objects of length 0 are silently dropped

name <- "Michael"
time_of_day <- "afternoon"
birthday <- FALSE

str_c(
  "Good ", time_of_day, " ", name, 
  if (birthday) " and HAPPY BIRTHDAY",
  "."
)

# to collapse a vector of strings into a single string, use collapse

str_c(c("x", "y", "z"), collapse = ", ")

```

### Subsetting Strings

```{r}

x <- c("Apple", "Banana", "Pear")

str_sub(x, 1, 3)

# negative numbers count backwards from end

str_sub(x, -3, -1)

#str_sub works even when the string is too short. It returns as much as possible

str_sub("a", 1, 5)

#str_sub can be used in assignment form to modify strings

str_sub(x, 1, 1) <- str_to_lower(str_sub(x, 1, 1))

```

### Locales

```{r}

# Turkish has 2 i': one with and one without a dot

str_to_upper("i")

str_to_upper("i", locale = "tr")

# Sorting is affected by locale as well

x <- c("apple", "eggplant", "banana")

str_sort(x, locale = "en")

str_sort(x, locale = "haw")

```

```{r}

# Whats the difference between paste() and paste0()? What stringr function are they equivalent to? 

paste("foo", "bar")
paste0("foo", "bar")

# paste includes spaces, paste0 does not

# close to paste0, different way of handling NA

str_c("foo", "bar")
str_c("foo", NA)

# paste treats NA as a character vector
paste("foo", NA)

# Difference between sep and collapse arguments in str_c? 

# sep inserts a new string between the parsed strings, whereas collapse treats the entire vector as a single string 

# Use str_length and str_sub to extract the middle character from a string. What will you do if the string has an even number of characters? 

strang_string <-"GilbertStraggqweqasdqgngDaNangThang"

x <- str_length(strang_string)

x

y <- ceiling(x/2)

str_sub(strang_string, y, y)

# What does str_wrap do? When might you want to use it? 

# str_wrap wraps strings into formatted paragraphs. This would be useful when reading in a formatted text that has lost its formatting when placed into a text file for program reading 

# What does str_trim() do? Whats the opposite of str_trim? 

# str_trim times whitespace from the start and end of a string. The opposite of str_trim would be str_pad 

str_trim(" abc  ", side = "left")
str_pad("abc", side = "both", width = 14)

# Write a function that turns a vector c("a", "b", "c") into the string a, b, and c. Think carefully about what it should do if given a vector of length 0, 1, or 2

str_commasep <- function(x, sep = ", ", last = ", and ") {
  if (length(x) > 1) {
    str_c(str_c(x[-length(x)], collapse = sep), 
          x[length(x)],
          sep = last)
  } else {
    x
  }
}

str_commasep("")
str_commasep("a")
str_commasep(c("a", "b"))
str_commasep(c("a", "b", "c"))

```

### Matching Patterns and Regular Expressions 

str_view()
str_view_all()

#### Basic Matches

```{r}
# match exact strings

x <- c("apple", "banana", "pear")

str_view(x, "an")

# match any character

str_view(x, ".a.")

# to create the regular expression with a predefined symbol, use \\

dot <- "\\."

writeLines(dot)

# tell R to look for an explicit

str_view(c("abc", "a.c", "bef"), "a\\.c")

# for a literal \

x <- "a\\b"

writeLines(x)

str_view(x, "\\\\")

# \ will escape the next character in the R string 
# \\ will resolve to \, which will escape the next character
# \\\ will resolve to a literal backslash in the reg exp, and the third will escape the next character. 

# \"\'\\\\

str_view("\"'\\", "\"'\\\\")

# What patterns will the regular expression \..\..\.. match? How would you represent it as a string? 

# any pattern where there is a . between each of the 3 characters. 

str_view(c(".a.b.c", ".a.b", "....."), c("\\..\\..\\.."))


```

### Anchors

^ to match the start of a string
$ to match the end of a string 

```{r}

x <- c("apple", "banana", "pear")
str_view(x, "^a")
str_view(x, "a$")

#mneumonic
# If you begin with power ^, you end up with money $

# to force a regexp to only match a complete string, anchor it with both ^ and $

x <- c("apple pie", "apple", "apple cake")
str_view(x, "apple")
str_view(x, "^apple$")
```

```{r}

# How would you match the literal string "$^$"?

x <- c("$^$", "$$$^^^", "^$^$^")

str_view(x, "^\\$\\^\\$$")

# Given the corpus of common words in stringr::words, create regexps that find all words that

# a) start with y

z <- stringr::words

str_view(z, "^y", match = TRUE)

# b) end with x

str_view(z, "x$", match = TRUE)

# c) are exactly 3 letters long 

str_view(z, "^...$", match = TRUE)

# d) have seven letters or more 

str_view(z, "^.......", match = TRUE)
```

### Character Classes and Alternatives

\d matches any digit
\s matches any whitespace
[abc] matches a, b, or c
[^abc] matches anything except a, b, or c

to use \d or \s, we need to escape the first \, so it is \\d or \\s

```{r}

# We can also use alternation

str_view(c("grey", "gray"), "gr(e|a)y")

```


```{r}

# Create regexps that find all words that

# a) start with a vowel
str_view(z, "^[aeiou]")

# b) only contain consonants
str_view(z, "^[^aeiou]+$", match = TRUE)

# c) end with ed, but not eed
str_view(z, "^ed$|[^e]ed$", match = TRUE)

# d) end with ing or ize
str_view(z, "ing$|ize$", match = TRUE)
str_view(z, "i(ng|ze)$", match = TRUE)

# empirically verify the rule i before e except after c
str_view(z, "(cei|[^c]ie)", match = TRUE)
str_view(z, "(cie|[^c]ei)", match = TRUE)

# with str_detect
sum(str_detect(z, "cei|[^c]ie"))
sum(str_detect(z, "(cie|[^c]ei)"))

# is q always followed by a u?

str_view(z, "q[^u]", match = TRUE)

# create a regexp that will match telephone numbers as commonly written in your country

x <- c("111-222-3333", "123-456-7890", "1800-64-giant")
str_view(x, "\\d\\d\\d-\\d\\d\\d-\\d\\d\\d\\d")

```

### Repetition

?: 0 or 1
+: 1 or more
*: 0 or more

```{r}

x <- "1888 is the longest year in roman numerals: MDCCCLXXXVIII"

# 0 or 1
str_view(x, "CC?")

# 1 or more
str_view(x, "CC+")


str_view(x, 'C[LX]+')
```

You can also specify the number of matches precisely: 

{n} - exactly n
{n,} - n or more
{,m} - at most m
{n,m} - between n and m

?: 0 or 1
+: 1 or more
*: 0 or more

```{r}

str_view(x, "C{2}")
str_view(x, "C{2,}")
str_view(x, "C{2,3}")

# by default, these matches are greedy - they will match the longest string possible. We can make them use lazy evaluation with ?

str_view(x, 'C{2,3}?')
str_view(x, 'C[LX]+?')

```

```{r}

# Describe the equivalents of ?, +, and * in {m,n} form
# ? is {,1}
# + is {1,}
# * is {0,}

# describe in words what these regexps match 
# a. ^.*$
#   begins with anything, 0 or more of something, ends
# b. "\\{.+\\}"
# Any string with curly braces surrounding at least one character
# c.\d{4}-\d{2}-\d{2}
# 4 digits - 2 digits - 2 digits, like 2018-01-14
# d. "\\\\{4}"
#   4 \'s

# create regexps to find all words that: 
# a. Start with 3 consonants
str_view(z, "^[^aeiou]{3}", match = TRUE)
# b. Have 3 or more vowels in a row
str_view(z, "[aeiou]{3,}", match = TRUE)
# c. Have 2 or more vowel consonant pairs in a row
str_view(z, "([aeiou][^aeiou]){2,}", match = TRUE)

```

### Grouping and Backreferences 

```{r}

# find all fruits with repeated pairs of letters
str_view(fruit, "(..)\\1", match = TRUE)

# Describe, in words, what these expressions will match: 
# a. (.)\1\1 - The same character appearing three times in a row
# b. "(.)(.)\\2\\1" - A pair of characters followed by the same pair in reverse. Ex: abba
# c. (..)\1 - 2 characters repeated. Ex: abab
# d. "(.).\\1.\\1" - An initial character, followed by anything, then orig, then anything, then orig - Ex: abaca
# e. "(.)(.)(.).*\\3\\2\\1" - c1c2c3 anything multiple times c3c2c1 - ex: abcafasdgasgasdgasddbca 

# Construct regexps to match words that 
# a. Start and end with the same character
str_view(z, "^(.).*\\1$", match = TRUE)

# b. Contain a repeated pair of letters 
str_view(z, "(..).*\\1", match = TRUE)
# more stringent
str_view(str_to_lower(z), "([a-z][a-z]).*\\1", match = TRUE)

# contain one letter repeated in at least 3 places 
str_view(z, "(.).*\\1.*\\1.*", match = TRUE)
```

### Tools

#### Detect Matches 

```{r}

x <- c("apple", "banana", "pear")

# str_detect 
str_detect(x, "e")

# How many common words start with t? 
sum(str_detect(words, "^t"))

# What proportion of common words end with a vowel? 
mean(str_detect(words, "[aeiou]$"))

# find words with no vowels
no_vowels_1 <- !str_detect(words, "[aeiou]")
# complex version of above
no_vowels_2 <- str_detect(words, "^[^aeiou]+$")
identical(no_vowels_1, no_vowels_2)

# use str_detect to select the elements that match a pattern

words[str_detect(words, "x$")]
str_subset(words, "x$")

# use filter instead

df <- tibble(
  words = words, 
  i = seq_along(word)
)

df %>% 
  filter(str_detect(words, "x$"))

# str_count tells you how many matches there are in a string 

x <- c("apple", "banana", "pear")
str_count(x, "a")

# on average, how many vowels per word? 
mean(str_count(words, "[aeiou]"))

# with mutate

df %>% 
  mutate(vowels = str_count(words, "[aeiou]"),
         consonants = str_count(words, "[^aeiou]"))

# in regexps, matches never overlap 
str_count("abababa", "aba")
str_view_all("abababa", "aba")
```

```{r}

# For each of the following challenges, try solving it by using both a single regular expression, and a combination of multiple str_detect calls 
# a. Find all words that start or end with x
str_view(words, "^x|x$", match = TRUE)
words[str_detect(words, "^x|x$")]

# split into parts
start_with_x <- str_detect(words, "^x")
end_with_x <- str_detect(words, "x$")
words[start_with_x | end_with_x]

# find all words that start with a vowel and end with a consonant
str_view(words, "^[aeiou].*[^aeiou]$", match = TRUE)

#str_subset
str_subset(words, "^[aeiou].*[^aeiou]$") %>% head()

# split into parts
start_with_vowel <- str_detect(words, "^[aeiou]")
end_with_consonant <- str_detect(words, "[^aeiou]$")
words[start_with_vowel & end_with_consonant]

# Are there any  words that contain at least one of each different vowel? 

# regexp of permutations
pattern <- cross(rerun(5, c("a", "e", "i", "o", "u")),
                   .filter = function(...) {
                     x <- as.character(unlist(list(...)))
                     length(x) != length(unique(x))
                   }) %>% 
  map_chr(~ str_c(unlist(.x), collapse = ".*")) %>% 
  str_c(collapse = "|")

str_subset(words, pattern)

words[str_detect(words, "a") &
      str_detect(words, "e") &
      str_detect(words, "i") &
      str_detect(words, "o") &
      str_detect(words, "u")]

# What word has the highest number of vowels? 
high_num_vowels <- str_count(words, "[aeiou]")
which.max(high_num_vowels)
words[48]

# What word has the highest proportion of vowels? 
high_prop_vowels <- high_num_vowels / str_length(words)
words[which(high_prop_vowels == max(high_prop_vowels))]

```

### Extract Matches

To extract the actual text of a match, we use str_extract()

```{r}
# harvard sentences. Designed to test VOIP
length(sentences)
head(sentences)

# find all sentences that contain a color
colors <- c("red", "orange", "yellow", "green", "blue", "purple")
# create vector with or statements
(color_match <- str_c(colors, collapse = "|"))

has_color <- str_subset(sentences, color_match)
matches <- str_extract(has_color, color_match)
head(matches)

# see more than one match
more <- sentences[str_count(sentences, color_match) > 1]
str_view_all(more, color_match)

# str_extract only shows the first encountered in a sentence
str_extract(more, color_match)

str_extract_all(more, color_match)

# If you use simplify = TRUE, str_extract_all will return a matrix with short matches expanded to the same length as the longest

str_extract_all(more, color_match, simplify = TRUE)

x <- c("a", "a b", "a b c")
str_extract_all(x, "[a-z]", simplify = TRUE)
```

```{r}

# From the harvard sentences data, extract: 

# a. The first words from each sentences
str_extract(sentences, "[a-zA-X]+")

# b. All words ending in ing
pattern <- "\\b[A-Za-z]+ing\\b"
sentences_with_ing <- str_detect(sentences, pattern)
unique(unlist(str_extract_all(sentences[sentences_with_ing], pattern))) %>% head()

# all plurals
unique(unlist(str_extract_all(sentences, "\\b[A-Za-z]{3,}s\\b")))

```

### Grouped Matches

```{r}
# extract nouns from sentences

# a sequence of 1 character that isn't a space after a or the
noun <- "(a|the) ([^ ]+)"

has_noun <- sentences %>% 
  str_subset(noun) %>% 
  head(10)

# str_extract gives us a complete match
has_noun %>% 
  str_extract(noun)

# str_match gives us each individual component
has_noun %>% 
  str_match(noun)

# if data is in a tibble, it is often easier to use tidyr::extract
tibble(sentence = sentences) %>% 
  tidyr::extract(
    sentence, c("article", "noun"), "(a|the) ([^ ]+)",
    remove = FALSE
  )

```

```{r}

# find all words that come after a number like "one", "two", "three", etc. Pull out both the number and the word
after_num <- "(one|two|three|four|five|six|seven|eight|nine|zero|ten|eleven|twelve) ([^ ]+)"

has_num <- sentences %>% 
  str_subset(after_num)

has_num %>% str_match(after_num)

# Find all contractions. Seperate out pieces before and after the apostrophe

contraction <- "([A-Za-z]+)'([A-Za-z]+)"
sentences %>% 
  `[`(str_detect(sentences, contraction)) %>% 
  str_extract(contraction)
```

### Replacing Matches

str_replace and str_replace_all

```{r}
# replace a pattern with a fixed string
x <- c("apple", "pear", "banana")
str_replace(x, "[aeiou]", "-")
str_replace_all(x, "[aeiou]", "-")

# str_replace_all with named vector
x <- c("1 house", "2 cars", "3 people")
str_replace_all(x, c("1" = "one", "2" = "two", "3" = "three"))

# Instead of replacing with a fixed string, you can use backreferences
sentences %>% 
  str_replace("([^ ]+) ([^ ]+) ([^ ]+)", "\\1 \\3 \\2") %>% 
  head()
```

```{r}
# Replace all forward slashes in a string with backslashes
backslashes <- str_replace_all(sentences, "\\/", "\\\\")
writeLines(backslashes)

# implement a simple version of str_to_lower() using replace_all()
lower <- str_replace_all(words, c("A" = "a", "B" = "b", "C" = "c","D" = "d","E" = "e","F" = "f","G" = "g","H" = "h","I" = "i","J" = "j","K" = "k","L" = "l","M" = "m","N" = "n","O" = "o","P" = "p","Q" = "q","R" = "r","S" = "s","T" = "t","U" = "u","V" = "v","W" = "w","X" = "x","Y" = "y","Z" = "z"))

# Switch the first and last letters in words. Which of those string are still words 

# make a vector of all the words with the first and last letters swapped
swapped <- str_replace_all(words, "^([A-Za-z])(.*)([a-z])$", "\\3\\2\\1")
# find out what swapped words are also in the original list
intersect(swapped, words)
```

### Splitting 

str_split() splits a word into pieces

```{r}
# split sentences into words
sentences %>% 
  head(5) %>% 
  str_split(" ")

# for length 1 vectors, split on the first element
"a|b|c|d" %>% 
  str_split("\\|") %>% 
  .[[1]]

# simplify = TRUE returns  a matrix
sentences %>% 
  head(5) %>% 
  str_split(" ", simplify = TRUE)

# request a maximum number of pieces
fields <- c("Name: Michael", "Country: USA", "Age: 26")
fields %>% 
  str_split(": ", n = 2, simplify = TRUE)

# split by character, line, sentence, and word boundaries
x <- "This is a sentence. This is another sentence."
str_view_all(x, boundary("word"))

str_split(x, " ")[[1]]

str_split(x, boundary("word"))[[1]]

```

```{r}

# split up a string like "apples, pears, and bananas" into individual components

fruits <- "apples, pears, and bananas"

str_split(fruits, boundary("word"))
str_split(fruits, ", +(and +)?")[[1]]


# why is it better to split up by boundary("word") than " "
# splitting by boundary("word") splits on punctuation and not just whitespace

# What does splitting with an empty string do? It splits the string into individual characters 
str_split(fruits, "")[[1]]
```

### Find Matches

str_locate() and str_locate_all() give the starting and ending positions of each match. You can use str_locate() to find the matching pattern and str_sub() to extract and/or modify them

### Other types of pattern

```{r}

# the regular call
str_view(fruit, "nana")
# is Shorthand for 
str_view(fruit, regex("nana"))

# ignore_case = TRUE
bananas <- c("banana", "Banana", "BANANA")
str_view(bananas, "banana")
str_view(bananas, regex("banana", ignore_case = TRUE))

# multiline = TRUE allows ^ and $ to match the start and end of each line rather than the start and end of the complete string 
x <- "Line 1\nLine 2\nLine 3"
str_extract_all(x, "^Line")[[1]]
str_extract_all(x, regex("^Line", multiline = TRUE))[[1]]

# comments = TRUE allows you to use comments and whitespace to make complex regular expressions more understandable. Spaces are ignored, as is everything after #. To match a literal space, you'll need to escape it \\ 
phone <- regex("
               \\(?     # optional opening parens
               (\\d{3}) # area code
               [)- ]?   # optional closing parens
               (\\d{3}) # another 3 numbers
               [ -]?    # optional space or dash
               (\\d{3}) # three more numbers
               ", comments = TRUE)
str_match("514-791-8141", phone)

# dotall = TRUE allows . to match everything, including \n

# There are 3 other functions that can be used instead of regex

# fixed() matches exactly the specified sequence of bytes. It ignores all special regexs and does its work at a low level making it about 3x as fast

microbenchmark::microbenchmark(
  fixed = str_detect(sentences, fixed("the")), 
  regex = str_detect(sentences, "the"),
  times = 20
)

# beware using fixed for non english data 
a1 <- "\u00e1"
a2 <- "a\u0301"
c(a1,a2)

a1 == a2

# we can use coll to change the human character comparison rules
str_detect(a1, fixed(a2))
str_detect(a1, coll(a2))

# coll() compares strings using standard collation rules. This makes it useful for case insensitive matching. It also changes for locale

# you can use boundary to match boundaries. You can also use it with the other functions

x <- "This is a sentence."
str_view_all(x, boundary("word"))
str_extract_all(x, boundary("word"))


```

```{r}

# How would you find all strings containing \ with regex() vs. fixed()?
str_subset(c("a\\b", "ab"), "\\\\")
str_subset(c("a\\b", "ab"), fixed("\\"))

# What are the five most common words in sentences? 

str_extract_all(sentences, boundary("word")) %>% 
  unlist() %>% 
  str_to_lower() %>% 
  tibble() %>% 
  set_names("word") %>% 
  group_by(word) %>% 
  count(sort = TRUE) %>% 
  head(5)

```

### Other Uses of Regular Expressions

```{r}

# apropos searches all objects available from the global environment
apropos("replace")

# dir() lists all the files in a directory 
head(dir(pattern = "\\.Rmd$"))
```

### Stringi

stringi has 234 functions, stringr has 42. stringr is the set of the most used functions in stringi

```{r}

# find all the stringi functions that
# a. Count the number of words: 
#stringi::stri_count_words(words)

# b. Find duplicated strings
#stringi::stri_duplicated()

# c Generate random text
# stringi::stri_rand_lipsum()
# stringi::stri_rand_strings()
# stringi::rand_shuffle 

# How do you control the language that stri_sort() uses for sorting? 
# Use the locale argument to the opts_collator argument

```


