---
title: "Untitled"
author: "Michael Rose"
date: "January 18, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(nycflights13)
```

## Date and Times with Lubridate

```{r}
today()
now()
```

### Creatng a date/time 

#### From Strings

```{r}
# to create a date
ymd("2017-01-31")
mdy("January 31st, 2017")
dmy("31-Jan-2017")

ymd(20170131)

# to create a datetime
ymd_hms("2017-01-31 20:11:59")
mdy_hm("01/31/2017 08:01")

# force a date time by supplying a timezone
ymd(20170131, tz = "UTC")
```

#### From Individual Components

```{r}
# select relevant columns
flights %>% 
  select(year, month, day, hour, minute)

# make_date() or make_datetime()
flights %>% 
  select(year, month, day, hour, minute) %>% 
  mutate(
    departure = make_datetime(year, month, day, hour, minute)
  )

# make datetime with modulus arithmetic to deal with time
make_datetime_100 <- function(year, month, day, time){
  make_datetime(year, month, day, time %/% 100, time %% 100)
}

flights_dt <- flights %>%
  filter(!is.na(dep_time), !is.na(arr_time)) %>% 
  mutate(
    dep_time = make_datetime_100(year, month, day, dep_time),
    arr_time = make_datetime_100(year, month, day, arr_time), 
    sched_dep_time = make_datetime_100(
      year, month, day, sched_dep_time
    ), 
    sched_arr_time = make_datetime_100(year, month, day, sched_arr_time)
  ) %>% 
  select(origin, dest, ends_with("delay"), ends_with("time"))

flights_dt

# view distribution of departure times across the year

flights_dt %>% 
  ggplot(aes(dep_time)) + 
  geom_freqpoly(binwidth = 86400)

# view the distribution of departure times within a single day

flights_dt %>% 
  filter(dep_time < ymd(20130102)) %>% 
  ggplot(aes(dep_time)) + 
  geom_freqpoly(binwidth = 600) # 600 s = 10 minutes
```

#### From other types

```{r}
# switch between a datetime and a date
as_datetime(today())
as_date(now())

# if the datetime is a numeric offset from the Unix Epoch (1970-01-01), then 
# if in seconds, use as_datetime
# if in days, use as_date
as_datetime(60 * 60 * 10)
as_date(365*10 + 2)

```

```{r}
# What happens when you parse a string that contains invalid dates?
ret <- ymd(c("2010-10-10", "bananas"))
print(class(ret))
ret

# What does the tzone argument to today() do? Why is it important? 
# a character vector specifying which time zone you would like to find the current date of. tzone defaults to the system time zone set on your computer. This is important for specifying which time zone you would like to use in your calculations, and also in analysis

# Use the appropriate lubridate function to parse each of the following dates: 

d1 <- "January 1, 2010"
d2 <- "2015-Mar-07"
d3 <- "06-Jun-2017"
d4 <- c("August 19 (2015)", "July 1 (2015)")
d5 <- "12/30/14" # Dec 30, 2014

mdy(d1)
ymd(d2)
dmy(d3)
mdy(d4)
mdy(d5)
```

### Date-Time Components

#### Getting Components

```{r}

datetime <- ymd_hms("2016-07-08 12:34:56")

year(datetime)
month(datetime)
mday(datetime)

yday(datetime)
wday(datetime)

# for month and wday, label = TRUE returns the abbreviated name of the month or day of the week
month(datetime, label = TRUE)
wday(datetime, label = TRUE)

# see distribution of flights by day

flights_dt %>% 
  mutate(wday = wday(dep_time, label = TRUE)) %>% 
  ggplot(aes(x = wday)) + 
  geom_bar()

# see distribution of flight delays by minute in an hour

flights_dt %>% 
  mutate(minute = minute(dep_time)) %>% 
  group_by(minute) %>% 
  summarize(
    avg_delay = mean(arr_delay, na.rm = TRUE),
    n = n()) %>% 
  ggplot(aes(minute, avg_delay)) + 
  geom_line()

# If we see a scheduled department time, there isn't such a strong pattern

sched_dep <- flights_dt %>% 
  mutate(minute = minute(sched_dep_time)) %>% 
  group_by(minute) %>% 
  summarize(
    avg_delay = mean(arr_delay, na.rm = TRUE), 
    n = n()
  )

ggplot(sched_dep, aes(minute, avg_delay)) + 
  geom_line()

# Like most patterns made by humans, there is a strong bias towards flights leaving at nice times 

ggplot(sched_dep, aes(minute, n)) + 
  geom_line()
```

#### Rounding 

```{r}
# plot the number of flights per week 
flights_dt %>% 
  count(week = floor_date(dep_time, "week")) %>% 
  ggplot(aes(week, n)) + 
  geom_line()

#we can also use round_date() and ceiling_date()
```
#### Setting Components

```{r}
datetime

# These are accessors for the components of the datetime

year(datetime) <- 2020
datetime

month(datetime) <- 01
datetime

hour(datetime) <- hour(datetime) + 1
datetime

# to set multiple values at once

update(datetime, year = 2020, month = 2, mday = 2, hour = 2)

# if the values are too big, they will roll over
ymd("2015-02-01") %>% 
  update(mday = 33)

ymd("2015-02-01") %>% 
  update(hour = 400)

# we can use update to show the distribution of flights across the course of the day for every day of the year

flights_dt %>% 
  mutate(dep_hour = update(dep_time, yday = 1)) %>% 
  ggplot(aes(dep_hour)) + 
    geom_freqpoly(binwidth = 300)

```

```{r}
# How does the distribution of flight times within a day change over the course of a year? 

# by month

flights_dt %>% 
  mutate(time = hour(dep_time) * 100 + minute(dep_time),
         mon = as.factor(month
                         (dep_time))) %>% 
  ggplot(aes(x = time, group = mon, color = mon)) + 
    geom_freqpoly(binwidth = 100)

# plot time against density 
flights_dt %>% 
  mutate(time = hour(dep_time) * 100 + minute(dep_time),
         mon = as.factor(month(dep_time))) %>% 
  ggplot(aes(x = time, y = ..density.., group = mon, color = mon)) + 
    geom_freqpoly(binwidth = 300)

# Compare dep_time, sched_dep_time, and dep_delay. Are they consistent? 

flights_dt %>% 
  mutate(dep_time_ = sched_dep_time + dep_delay * 60) %>% 
  filter(dep_time_ != dep_time) %>% 
  select(dep_time_, dep_time, sched_dep_time, dep_delay)

# There exist some discrepancies, such as differing dates and times 

# Compare air_time with the duration between the departure and arrival. Explain your findings

flights_dt %>% 
  mutate(flight_duration = as.numeric(arr_time - dep_time),
         air_time_mins = air_time,
         diff = flight_duration - air_time_mins) %>% 
  select(origin, dest, flight_duration, air_time_mins, diff)

# How does the average delay time change over the course of a day? Should you use dep_time or sched_dep_time? 

# We should use sched_dep_time since it is the time that attendees seek, whereas dep_time will have a lag throughout the day due to flights getting pushed back

# average delay time change over the course of a day

flights_dt %>% 
  mutate(sched_dep_hour = hour(sched_dep_time)) %>% 
  group_by(sched_dep_hour) %>% 
  summarize(dep_delay = mean(dep_delay)) %>% 
  ggplot(aes(y = dep_delay, x = sched_dep_hour)) + 
    geom_point() + geom_smooth()

# On what day of the week should you leave if you want to minimize the chance of a delay? 

flights_dt %>% 
  mutate(dow = wday(sched_dep_time)) %>% 
  group_by(dow) %>% 
  summarize(dep_delay = mean(dep_delay),
            arr_delay = mean(arr_delay, na.rm = TRUE))

# What makes the distribution of diamonds$carat and flights$sched_dep_time similar? 

ggplot(diamonds, aes(x = carat)) + 
  geom_density()

# In both carat and sched_dep_time, there are lots of numbers at human values. In carat, they are things like 0, 1/3, 1/2, 2/3 and in sched_dep_time they are things like 00, 30, X0, X5

ggplot(diamonds, aes(x = carat %% 1 * 100)) + 
  geom_histogram(binwidth = 1)

ggplot(flights_dt, aes(x = minute(sched_dep_time))) + 
  geom_histogram(binwidth = 1)

# Confirm my hypothesis that the early departures of flights in minutes 20 - 30, and 50 - 60 are caused by scheduled flights that leave early. 

# At the minute level, there doesn't appear to be anything 
flights_dt %>% 
  mutate(early = dep_delay < 0,
         minute = minute(sched_dep_time)) %>% 
  group_by(minute) %>% 
  summarize(early = mean(early)) %>% 
  ggplot(aes(x = minute, y = early)) + 
    geom_point()

# grouped in 10 minute intervals

flights_dt %>% 
  mutate(early = dep_delay < 0,
         minute = minute(sched_dep_time) %% 10) %>% 
  group_by(minute) %>% 
  summarize(early = mean(early)) %>% 
  ggplot(aes(x = minute, y = early)) + 
    geom_point()
```

### Time Spans

Durations represent an exact number of seconds
Periods represent human units like weeks and months
Intervals represent a starting and ending point

#### Durations 

```{r}

# In R, when 2 dates are subtracted we get a difftime object

m_age <- today() - ymd(19911202)
m_age

k_age <- today() - ymd(19960111)
k_age

# alternative in seconds with an approx age in years
as.duration(m_age)
as.duration(k_age)

# constructors
dseconds(15)
dminutes(10)
dhours(c(12, 24))
ddays(0:5)
dweeks(3)
dyears(1)

# durations can be added and multiplied
2 * dyears(1)
dyears(1) + dweeks(12) + dhours(15)

# we can add and subtract durations to and from days
tomorrow <- today() + ddays(1)
tomorrow

last_year <- today() - dyears(1)
last_year

# due to exact numbers of seconds, sometimes we get an unexpected result
one_pm <- ymd_hms(
  "2016-03-12 13:00:00",
  tz = "America/New_York"
)

one_pm
one_pm + ddays(1)

# This changed from 1 to 2 pm due to daylight savings time!
```

#### Periods

Periods don't have a fixed length in seconds, and instead operate like human times, like days and months.

```{r}
one_pm
one_pm + days(1)

# constructor functions
seconds(15)
minutes(10)
hours(c(12, 24))
days(7)
months(1:6)
weeks(3)
years(1)

# you can add and multiply periods
10 * (months(6) + days(1))
days(50) + hours(25) + minutes(2)

# add to dates

# a leap year
ymd("2016-01-01") + dyears(1)
ymd("2016-01-01") + years(1)

# daylight savings time
one_pm + ddays(1)
one_pm + days(1)

# using periods to fix an oddity in flight dates

# find overnight flights
flights_dt %>% 
  filter(arr_time < dep_time)

# fix this error by adding a day to the arr_times
flights_dt <- flights_dt %>% 
  mutate(
    overnight = arr_time < dep_time,
    arr_time = arr_time + days(overnight * 1),
    sched_arr_time = sched_arr_time + days(overnight * 1)
  )

# check to make sure it worked
flights_dt %>% 
  filter(overnight, arr_time < dep_time)

```

#### Intervals

```{r}
# gives a warning when dates are inconsistent, such as leap years not being 365 days

years(1) / days(365)

# an interval is more accurate

next_year <- today() + years(1)
(today() %--% next_year) / ddays(1)

# to see how many periods fall into an interval, use integer division
(today() %--% next_year) %/% days(1)
```

#### Summary

If you only care about a physical time, use a duration 
if you need to add human times, use a period
If you need to figure out how long a span is in human units, use an interval

```{r}
# Why is there months(), but no dmonths()?
# There is no direct value of months to seconds because months can be 28,29,30 or 31 days long

# Explain days(overnight * 1) to someone who has just started learning R. How does it work? 
# T/F values turn into 1 or 0. So 1*1 = 1, and adds 1 and 0*1 = 0 and adds 0

# create a vector of dates giving the first day of every month in 2015. Create a vecotr of dates giving the first day of every month in the current year 

# 2015
ymd(20150101) + months(0:11)

# This year
floor_date(today(), unit = "year") + months(0:11)


# write a function that, given your birthday as a dte, returns how old you are in years

age <- function(bday) {
   bday <- ymd(bday)
   (bday %--% today()) %/% years(1)
}

age(19911202)

# Why can't today() %--% (today() + years(1)) / months(1) work? 
(today() %--% (today() + years(1))) / months(1)
# it does work 

```

### Time Zones

```{r}
# check current timezone

Sys.timezone()

# complete list of all timezones
length(OlsonNames())

head(OlsonNames())

# these 3 objects represent the same time 

x1 <- ymd_hms("2015-06-01 12:00:00", tz = "America/New_York")
x2 <- ymd_hms("2015-06-01 18:00:00", tz = "Europe/Copenhagen")
x3 <- ymd_hms("2015-06-02 04:00:00", tz = "Pacific/Auckland")

# verify with subtraction
x1 - x2
x1 - x3

# Operations that combine datetimes, such as c() drop the time zone and display the users timezone

x4 <- c(x1, x2, x3)
x4

# changing timezones

x4a <- with_tz(x4, tzone = "Australia/Lord_Howe")
x4a

x4a - x4

# change the underlying instant in time, forcing the change in timezone due to an error in tz
(x4b <- force_tz(x4, tzone = "Australia/Lord_Howe"))
```

