---
title: "Untitled"
author: "Michael Rose"
date: "January 21, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lubridate)
library(tidyverse)
```

### When should you write a function? 

```{r}
# functions should be written whenever code is going to be pasted more than twice

df <- tibble::tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

# code to be turned into function for inputs a, b, c, d

(df$a - min(df$a, na.rm = TRUE)) / (max(df$a, na.rm = TRUE) - min(df$a, na.rm = TRUE))

# break down

# input
x <- df$a
# plugging x in
(x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))


# we are computing the range 3 times
rng <- range(x, na.rm = TRUE)
# plugging it in
(x - rng[1]) / (rng[2] - rng[1])

# turn it into a function
rescale01 <- function(x){
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}

# test it
rescale01(x)
rescale01(c(0,5,10))

# simplify original problem with function
df$a <- rescale01(df$a)
df$b <- rescale01(df$b)
df$c <- rescale01(df$c)
df$d <- rescale01(df$d)

# if rescale fails (like with infinite values), we only need to modify the function
x <- c(1:10, Inf)
rescale01(x)

# fix function
rescale01 <- function(x){
  rng <- range(x, na.rm = TRUE, finite = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}

# DRY - Do Not Repeat Yourself
```

```{r}
# why is TRUE not a parameter to rescale01? What would happen if x contained a single missing value, and na.rm was false? 

# if na.rm was false, then the function would return NA. TRUE is not a parameter because we generally always want to remove NA and it is not a part of the functions that is generally changed

rescale01_alt <- function(x, finite = TRUE){
  rng <- range(x, na.rm = finite, finite = finite)
  (x - rng[1])/(rng[2]-rng[1])
}

rescale01_alt(c(NA, 1:5), finite = FALSE)
rescale01_alt(c(NA, 1:5), finite = TRUE)

# In the second variant of rescale01, infinite values are left unchanged. Rewrite rescale01 so that -Inf is mapped to 0, and Inf is mapped to 1

rescale01 <- function(x){
  rng <- range(x, na.rm = TRUE, finite = TRUE)
  y <- (x - rng[1])/(rng[2] - rng[1])
  y[y==Inf] <- 1
  y[y==-Inf] <- 0
  y
}

rescale01(c(Inf, -Inf, 0:5, NA))

# Practice turning the following code snippets into functions 

mean(is.na(x))

prop_na <- function(x){
  mean(is.na(x))
}

prop_na(c(NA, 0, NA, 0, NA))

x / sum(x, na.rm = TRUE)

weights <- function(x){
  x / sum(x, na.rm = TRUE)
}

(y <- weights(0:5))

# This function calculates the coefficient of variation (assuming that x can only take non-negative values). The coefficient of variation is the standard deviation divided by the mean: 

sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE)

coef_of_var <- function(x){
  sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE)
}

coef_of_var(runif(10))

# do https://nicercode.github.io/intro/writing-functions.html

herb <- read_csv("seed_root_herbivores.csv")

mean(herb$Height)
var(herb$Height)
# sample size
length(herb$Height)

# standard error
sqrt(var(herb$Height) / length(herb$Height))

# standard error of dry weight
sqrt(var(herb$Weight) / length(herb$Weight))

# syntax

standard.error <- function(x){
  sqrt(var(x) / length(x))
}

standard.error(herb$Height)
standard.error(herb$Weight)

# defining variables within functions

standard.error <- function(x){
  v <- var(x)
  n <- length(x)
  sqrt(v/n)
}

standard.error(herb$Height)

# Variance function

x <- herb$Height
n <- length(x)
(1/(n-1))
m <- mean(x)
x - m
(x - m)^2
sum((x-m)^2)

# var
(1/(n-1)) * sum((x-m)^2)
var(x)

# clean up global environment
rm(n, x, m)

variance <- function(x){
  n <- length(x)
  m <- mean(x)
  (1/(n-1))*sum((x-m)^2)
}

variance(herb$Height)
var(herb$Height)

variance(herb$Weight)
var(herb$Weight)

# an aside on floating point comparisons

variance(herb$Height) == var(herb$Height)

variance(herb$Height) - var(herb$Height)

sqrt(2) * sqrt(2)

sqrt(2) * sqrt(2) - 2

# don't use ==, use abs(x-y) < eps

skewness <- function(x){
  m <- mean(x)
  n <- length(x)
  v <- var(x)
  third.moment <- (1/(n-2)) * sum((x-m)^3)
  third.moment / var(x)^(3/2)
}

skewness(herb$Height)
skewness(herb$Weight)

variance <- function(x){
  # remove missing values
  x <- x[!is.na(x)]
  n <- length(x)
  m <- mean(x)
  sq_err <- (x-m)^2
  sum(sq_err) / (n-1)
}

var(1:10)
variance(1:10)

skewness <- function(x){
  x <- x[!is.na(x)]
  n <- length(x)
  m <- mean(x)
  m3 <- sum((x-m)^3) / n
  s3 <- sqrt(sum((x-m)^2) / (n-1))
  m3 / s3
}

skewness(rgamma(10, 1, 1))

# Write both_na(), a function that takes two vectors of the same length and returns the number of positions that have an NA in both vectors

both_na <- function(x, y){
  vec1 <- sum(is.na(x))
  vec2 <- sum(is.na(y))
  vec1 + vec2
}

que1 <- c(NA, "a", "b", "c", NA, NA)
que2 <- c(NA, "a", "b", "c", NA, "d")

both_na(que1, que2)

both_na <- function(x, y){
  sum(is.na(x) & is.na(y))
}

both_na(que1, que2)

# What do the following functions do? Why are they useful even though they are short? 

# this checks whether x is a directory
is_directory <- function(x){
  file.info(x)$isdir
}

# this checks whether the file is accessible 
is_readable <- function(x){file.access(x, 4) == 0}

```

### Functions are for humans and computers 

```{r}

# read the source code for each of the following 3 functions, puzzle out what they do, and then brainstorm better names
f1 <- function(string, prefix){
  substr(string, 1, nchar(prefix)) == prefix
}

# f1 checks whether the string of names has a common prefix. A better name would be same_prefix 

f2 <- function(x){
  if (length(x) <= 1)
    return(NULL)
  x[-length(x)]
}

# f2 checks if the length of x is <= 1, and if not it removes the last element. A better name would be drop_last

f3 <- function(x, y){
  rep(y, length.out = length(x))
}

# f3 replicates the element y exactly length(x) times. A better name would be num_reps

f3(c(1, 2, 3, 4, 5), "z")

# Compare and contrast rnorm() and MASS::mvrnorm(). How could you make them more consistent? 

# rnorm samples from the univariate normal distribution and mvrnorm samples from the multivariate normal distribution. They could be made more consistent by having the same name and using method overloading to read in parameters and decide whether to use the uni or multivariate norm function

# Make a case for why norm_r(), norm_d(), etc, would be better than rnorm(), dnorm(). Make a case for the opposite

# If we grouped by norm, it would show that the functions are all based around the normal distribution. If we grouped by r, d, we know every r function samples from the distribution following (such as rnorm, runif, rexp), and every d function calculate the probability density (dnorm, dbinom, dexp, dunif)
```

### Conditional Execution

```{r}
# Return a logical vector describing whether or not each element of a vector is named 

has_name <- function(x){
  nms <- names(x)
  if (is.null(nms)){
    rep(FALSE, length(x))
  } else {
    !is.na(nms) & nms != ""
  }
}

```

#### Conditions

if given a long chain of if statements, a useful technique is the switch function

ex: 

function(x, y, op){
  switch(op,
    plus = x + y, 
    minus = x - y,
    times = x * y,
    divide = x / y,
    stop("Unknown op!"))
}

### Code Style

```{r}
# good

if (y < 0 && debug){
  message("Y is negative")
}

if (y == 0){
  log(x)
} else {
  y ^ x
}

```

```{r}
# what is the difference between if and ifelse? 
# The keyword if checks one condition, whereas ifelse checks multiple conditions

#x <- 3

#ifelse(x %%2 == 0){
#  TRUE
#  FALSE
#}


# write a greeting function tht says good morning, afternoon, or evening depending on the time of day
good_x <- function(time = hour(lubridate::now())){
  if (time < 12){
    print('good morning')
  }
  else if (time < 18){
    print('good afternoon')
  }
  else {
      print('good evening')
  }
}

good_x()

# implement a fizzbuzz function. 

fizzbuzz <- function(n){
  if (n %% 3 == 0 & n%%5 == 0){
    print('fizzbuzz')
  }
  if (n%%3 == 0){
    print('fizz')
  }
  if (n%%5 == 0){
    print('buzz')
  }
}


fizzbuzz(15)


fizzbuzz <- function(x) {
  stopifnot(length(x) == 1)
  stopifnot(is.numeric(x))
  # this could be made more efficient by minimizing the
  # number of tests
  if (!(x %% 3) & !(x %% 5)) {
    print("fizzbuzz")
  } else if (!(x %% 3)) {
    print("fizz")
  } else if (!(x %% 5)) {
    print("buzz")
  }
}
fizzbuzz(6)
fizzbuzz(10)
fizzbuzz(15)
fizzbuzz(2)

# weird

# how could you use cut() to simplify this set of nested if else statements? 

if (temp <= 0){
  "freezing"
} else if (temp <= 10){
  "cold"
} else if (temp <= 20) {
  "cool"
} else if (temp <= 30) {
  "warm"
} else {
  "hot"
}

temp <- seq(-10, 50, by = 5)
cut(temp, c(-Inf, 0, 10, 20, 30, Inf), right = TRUE, 
    labels = c("Freezing", "Cold", "Cool", "Warm", "Hot"))

# to have intervals open on the left (using <), change te argument to right = FALSE

temp <- seq(-10, 50, by = 5)
cut(temp, c(-Inf, 0, 10, 20, 30, Inf), right = FALSE, 
    labels = c("Freezing", "Cold", "Cool", "Warm", "Hot"))

# two advantages of cut: 
# 1. It works on vectors, whereas if only works on a single value 
# 2. To change comparisons, we only need to change the right argument

# What happens if you use switch() with numeric values? 

x <- '3'
switch(x,
       '1' = "a",
       '2' = "b",
       '3' = "c"
       )

# it works if we treat the numeric value as a string, but not otherwise

# What does this switch() call do? What happens if x is 'e'?

x <- 'a'
switch(x,
       a = ,
       b = 'ab',
       c = ,
       d = 'cd'
       )

# it returns ab for a or b, cd for c or d and nothing otherwise

```

### Function Arguments

```{r}
# Compute confidence interval around mean using normal approx
mean_ci <- function(x, conf = 0.95){
  se <- sd(x) / sqrt(length(x))
  alpha <- 1 - conf 
  mean(x) + se * qnorm(c(alpha / 2, 1 - alpha / 2))
}

x <- runif(100)

mean_ci(x)

mean_ci(x, conf = 0.99)

# weighted summary statistics

wt_mean <- function(x, w){
  sum(x * w) / sum(x)
}

wt_var <- function(x, w){
  mu <- wt_mean(x, w)
  sum(w * (x - mu) ^ 2) / sum(w) 
}

wt_sd <- function(x, w){
  sqrt(wt_var(x, w))
}

# what happens if x and w are not the same length? 

wt_mean(1:6, 1:3)

# R has vector recycling rules. Therefore its important to check preconditions and throw an error if they are not true

wt_mean <- function(x, w){
  if (length(x) != length(w)){
    stop("'x' and 'w' must be same length", call. = FALSE)
  }
  sum(w * x)/sum(x)
}

wt_mean(1:6, 1:3)

# There is a tradeoff between how much time is spent making a function robust vs how long is spent writing it

wt_mean <- function(x, w, na.rm = FALSE){
  if (!is.logical(na.rm)){
    stop('na.rm must be logical')
  }
  if (length(na.rm) != 1){
    stop('na.rm must be length 1')
  }
  if (length(x) != length(w)){
    stop('x and w must be same length')
  }
  
  if (na.rm){
    miss <- is.na(x) | is.na(w)
    x <- x[!miss]
    w <- w[!miss]
  }
  sum(w * x) / sum(x)
}

# stopifnot checks if each argument is true and throws an error message if not

wt_mean <- function(x, w, na.rm = FALSE){
  stopifnot(is.logical(na.rm), length(na.rm == 1))
  stopifnot(length(x) == length(w))
  
  if (na.rm){
    miss <- is.na(x) | is.na(w)
    x <- x[!miss]
    w <- w[!miss]
  }
  sum(w * x) / sum(x)
}

wt_mean(1:6, 6:1, na.rm = 'foo')

```

### Dot-dot-dot(...)

```{r}
# ... captures any arguments that aren't otherwise matched, allowing functions like
sum(1,2,3,4,5,6,7,8,9,10)

stringr::str_c('a', 'b', 'c')

# this is a useful catch all if your function wraps around another function

commas <- function(...) stringr::str_c(..., collapse = ', ')
commas(letters[1:10])

rule <- function(..., pad = '-'){
  title <- paste0(...)
  width <- getOption('width') - nchar(title) - 5
  cat(title, " ", stringr::str_dup(pad, width), "\n", sep = "")
}

rule("Important Output")
```

```{r}

# What does commas(letters, collapse = "-") do? Why?
commas(letters, collapse = "-")
# throws an error. Collapse is passed to str_c as part of ..., so it tries to run str_c(letters, collapse="-", collapse=", ")

# It'd be nice if you could supply multiple characters to the pad argument, eg. rule("Title", pad = "-+"). Why doesn't this currently work? How could you fix it? 

rule <- function(..., pad = "-"){
  title <- paste0(...)
  width <- getOption("width") - nchar(title) - 5
  cat(title, " ", stringr::str_dup(pad, width), '\n', sep = "")
}

rule("Important Output")
rule("Important Output", pad = "-+")

# this does not work because the width is preset

rule <- function(..., pad = "-") {
  title <- paste0(...)
  width <- getOption("width") - nchar(title) - 5
  padchar <- nchar(pad)
  cat(title, " ", 
      stringr::str_dup(pad, width %/% padchar), 
      # if not multiple, fill in the remaining chars
      stringr::str_sub(pad, 1, width %% padchar), 
      "\n", sep = "")
}

rule("Important Output")
rule("Important Output", pad = "+-")
rule("Important Output", pad = "123")

# What does the trim argument to mean do? When might you use it? 

# trim removes a % of the extrema from the sample when calculating the mean. This is useful for calculating a mean that is robust to outliers

# The default value for the method argument to cor() is c("pearson", "kendall", "spearman"). What does that mean? What value is used by default? 

# It means that cor() can use the methods pearson, kendall, or spearman. Pearson is used by default 

```

### Return Values 

```{r}

# a common early return reason is empty inputs

comp_func <- function(x, y, z){
  if (length(x) == 0 || length(y) == 0){
    return(0)
  }
  else{
    # rest of code
  }
}


```

### Writing Pipeable Functions

2 main types: 

- transformation: a clear 'primary object' that is passed in as the first arg and a modified version is returned
- side effect: do an action, like save a file, draw a plot, etc

```{r}
# print number of missing values in a data frame

show_missings <- function(df){
  n <- sum(is.na(df))
  cat("Missing values: ", n, "\n", sep = "")
  # make sure the df doesn't get printed out
  invisible(df)
}

x <- show_missings(mtcars)

# the df is still there, its just not printed out

class(x)

dim(x)

mtcars %>% 
  show_missings() %>% 
  mutate(mpg = ifelse(mpg < 20, NA, mpg)) %>% 
  show_missings() 

```

### Environment

```{r}

```

