---
title: "Untitled"
author: "Michael Rose"
date: "February 3, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

### Vector Basics

Every vector has 2 key properties: 

```{r}
# type
typeof(letters)
typeof(1:10)

# length
x <- list("a", "b", 1:10)
length(x)

# vectors can also contain arbitrary additional metadata in the form of attributes. These create augmented vectors. 
# Factors are built on top of integer vectors
# dates and date-times are built on top of numeric vectors
# data frames and tibbles are built on top of lists
```

### Important Types of Atomic Vector

#### Logical, Integer, Double, and Character

```{r}

1:10%%3 == 0
# create with c
c(TRUE, FALSE, FALSE, NA)

# Numeric

# in R, numbers are doubles by default. To coerce to int, place an L after
typeof(1)
typeof(1L)
1.5L


# R uses a global string pool, meaning that each unique sting is only stored in memory once, and every use of the string points to that representation. 

x <- "This is a reasonably long string"
pryr::object_size(x)

y <- rep(x, 1000)
pryr::object_size(y)

```

```{r}
# Describe the difference between is.finite() and !is.infinite()

# Try the function on a numeric vector that includes a number and the five special values NA, NaN, Inf, -Inf

x <- c(0, NA, NaN, Inf, -Inf)
is.finite(x)
!is.infinite(x)

# Read the source code to dplyr::near
#> dplyr::near
#function (x, y, tol = .Machine$double.eps^0.5) 
#{
#    abs(x - y) < tol
#}
#<environment: namespace:dplyr>

# It works by creating a tolerance interval epsilon and seeing if the absolute value of x - y is within that interval 

# A logical vector can take three possible values. How many possible values can integer take? double? 

# From ?.Machine

# The algorithm is based on Cody's (1988) subroutine MACHAR. As all current implementations of R use 32-bit integers and use IEC 60559 floating-point (double precision) arithmetic, all but three of the last four values are the same for almost all R builds.

#Note that on most platforms smaller positive values than .Machine$double.xmin can occur. On a typical R platform the smallest positive double is about 5e-324.

# Brainstorm at least four functions that allow you to convert a double to an integer. How do they differ? 

# Truncate to 1 digit, floor, ceiling, any arbitrary rounding function 

# Rounding is important! 

# https://en.wikipedia.org/wiki/Vancouver_Stock_Exchange 
# The history of the exchange's index provides a standard case example of large errors arising from seemingly innocuous floating point calculations. In January 1982 the index was initialized at 1000 and subsequently updated and truncated to three decimal places on each trade. Such a thing was done about 3000 times each day. The accumulated truncations led to an erroneous loss of around 25 points per month. Over the weekend of November 25–28, 1983, the error was corrected, raising the value of the index from its Friday closing figure of 524.811 to 1098.892.

# What functions from the readr package allow you to turn a string into logical, integer, and double vector? 

# parse_logical
parse_logical(c("TRUE", "TRUE", "1", "0", "true", "t", "NA"))

# parse_integer
parse_integer(c("12345", "5678", "NA"))

# parse_double
parse_double(c("1.0", "2.0", "1000", "NA"))
```

### Using Atomic Vectors

#### Coercion

Explicit Coercion happens when you call a function like as.numeric or as.logical. Whenever used, we should check if the function could be called upstream. 

Implicit Coercion happens when you use a vector in a specific context that expects a certain type of vector. Ex: logical vector with a numeric summary function 

```{r}
x <- sample(20, 100, replace = TRUE)
y <- x > 10

# sum is number of trues, mean is proportion of trues
sum(y)
mean(y)

# When creating a vector with c, the most complex type wins
typeof(c(1L, TRUE))
typeof(c(1L, 1.5))
typeof(c(1.5, "a"))

```

#### Scalars and Recycling Rules

R implicitly recycles vectors. That means that if we have 2 vectors, the shorter one is repeated until it is the same length as the longer one. 

```{r}
sample(10) + 100

runif(10) > 0.5

# Add 2 vectors of different length

1:10 + 1:2

# when shorter not a multiple of longer
1:10 + 1:3

# by default, tidyverse throws an error
tibble(x = 1:4, y = 1:2)

# force recycle 
tibble(x = 1:4, y = rep(1:2, 2))
tibble(x = 1:4, y = rep(1:2, each = 2))

# all types of vectors can be named
# during
c(x = 1, y = 2, z = 3)

# afterwards
set_names(1:3, c("a", "b", "c"))


```


#### Subsetting 

```{r}
x <- c("one", "two", "three", "four", "five")
x[c(3,2,5)]

# output can be longer than input
x[c(1,1,2,5,5,5,3)]

# negative values drop the elements at specific positions
x[c(-1, -3, -5)]

# error to mix positives and negatives
x[c(1, -1)]

x[0]

x <- c(10, 3, NA, 5, 8, 1, NA)
x[!is.na(x)]

x[x%%2 == 0]

# for named vectors
x <- c(abc = 1, def = 2, xyz = 5)
x[c('xyz', 'def')]

# select first row and all columns
x[1, ]

# select all rows and all columns except the first
x[, -1]

```


```{r}

# What does mean(is.na(x)) tell you about vector x? What about sum(!is.finite(x))? 

# mean(is.na(x)) shows the proportion of NAs in the vector
# sum(!is.finite(x)) shows how many NAs, NaNs, or +-infs there are in the vector

# What does is.vector() test for? Why does is.atomic() not agree wit the definition of atomic vectors? 

# is.vector checks that an object has no other attributes other than names. Any object with attributes besides names doesn't work 

# is.atomic checks whether an object is one of the atomic types, like logical, int, double, complex, character, or raw

# Compare and contrast setNames() and purrr::set_names()

purrr::set_names

setNames

# purrr gives it a whole bunch of defense for the inputs

# create functions that take a vector as an input and return: 

# a. The last value. 

last_val <- function(v){
  return(v[length(v)])
}

last_val(x)

last_value <- function(x){
  # check for case with no length
  if (length(x)) {
    # use [[ as suggested because it returns 1 element
    x[[length(x)]]
  } else {
    x
  }
}

last_value(x)
last_value(numeric())
last_value(1)
last_value(1:10)

# b. The elements at even numbered positions

even_pos <- function(v){
  # for for length
  if (length(x)){
    x[x%%2 == 0]
  } else {
    x
  }
}

even_pos(x)

even_indices <- function(x){
  if (length(x)){
    x[seq_along(x) %% 2 == 0]
  } else {
    x
  }
}

even_indices(x)
even_indices(numeric())
even_indices(1)
even_indices(1:10)
even_indices(letters)

# c. Even element except the last value 

not_last <- function(v){
  if (length(v) != 0){
    v[-length(v)]
  } else{
    v
  }
}

not_last(x)

not_last_2 <- function(x){
  if (length(x)){
    x[-length(x)]
  } else {
    x
  }
}

not_last_2(1:5)

# Only even numbers and no missing values

evens_no_na <- function(v){
  if (length(v) > 0){
    v[v%%2 == 0 & !is.na(v)]
  } else {
    v
  }
}

q <- c(1:5, NA, 25:35, NaN)

evens_no_na(q)

# Why is x[-which(x > 0)] not the same as x[x <= 0]?

# They treat missing values differently

q[-which(q > 0)]
-which(q > 0)
q[q <= 0]
q <= 0

# What happens when you subset with a positive integer that's bigger than the length of the vector? What happens when you subset with a name that doesn't exist? 

(1:10)[11:12]

c(a = 1, b = 2)[['c']]
```

### Recursive Vectors (Lists) 

```{r}
x <- list(1, 2, 3)
x

# str focuses on the -str-ucture, not the contents
str(x)

x_named <- list(a = 1, b = 2, c = 3)
str(x_named)

# lists can contain a mix of objects

y <- list("a", 1L, 1.5, TRUE)
str(y)

# lists within lists! 

z <- list(list(1, 2), list(3, 4))

str(z)

# subsetting 

a <- list(a = 1:3, b = "a string", c = pi, d = list(-1, -5))

# [ extracts a sublist, returning a list

str(a[1:2])

# [[]] extracts a single component from a list. It removes a level of hierarchy

str(y[[1]])
str(y[[4]])

# $ is shorthand for extracting named elements of a list 
a$a
a[['a']]


```

```{r}
# What happens if you subset a tibble as if you're subsetting a list? What are the key differences between a list and a tibble? 

x <- tibble(a = 1:2, b = 3:4)
x[['a']]
x$a
x[1]
x[1, ]

# They work the same, because a tibble can be thought of as a list of columns. The key difference between the list and tibble (data frame ) is the restriction that each of the columns (lists) must be of the same length


```

### Attributes 

We can think of attributes as a named list of vectors that can be attached to any object

```{r}
x <- 1:10
attr(x, "greeting")
attr(x, "greeting") <- "Hi!"
attr(x, "farewell") <- "Bye!"

attributes(x)

# names are used to name the elements of a vector
# Dimensions or dims make a vector behave like a matrix or array
# class is used to implement the S3 object oriented system 

# list all the methods for a generic
methods('as.Date')
```

### Augmented Vectors

```{r}
# Factors
# represent categorical data that can take a fixed set of possible values
x <- factor(c("ab", "cd", "ab"), levels = c("ab", "cd", "ef"))
typeof(x)
attributes(x)

# Dates and Date-Times
# dates in R are numeric vectors that represent the number of days since January 1st, 1970

x <- as.Date('1971-01-01')
unclass(x)
typeof(x)
attributes(x)

x <- lubridate::ymd_hm("1970-01-01 01:00")
unclass(x)
typeof(x)
attributes(x)

# change timezone
attr(x, 'tzone') <- "US/Pacific"
x

attr(x, 'tzone') <- "US/Eastern"
x

# POSIXlt

y <- as.POSIXlt(x)
typeof(y)
attributes(y)

# tibbles

tb <- tibble(x = 1:5, y = 5:1)
typeof(tb)
attributes(tb)

# traditional data frames

df <- data.frame(x = 1:5, y = 5:1)
typeof(df)
attributes(df)
```

```{r}

# hms::hms returns an object of class and prints the time in HMS format
x <- hms::hms(3600)
class(x)
x

# double
typeof(x)

# units and class
attributes(x)

# try to make a tibble that has columns with different lengths. What happens? 

tibble(x = 1, y = 1:5)

# the shorter vector is recycled

tibble(x = 1:3, y = 1:5)

# error when not scalar multiples

# Based off the previous definition, is it OK to have a list as a column of a tibble? 


tibble(x = 1:3, y = list("a", 2, list(1:5)))
```
