---
title: "Chapter 1_R4DS"
author: "Michael Rose"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

A data frame with 234 rows and 11 variables

manufacturer
model
model name

displ
engine displacement, in litres

year
year of manufacture

cyl
number of cylinders

trans
type of transmission

drv
f = front-wheel drive, r = rear wheel drive, 4 = 4wd

cty
city miles per gallon

hwy
highway miles per gallon

fl
fuel type

class
"type" of car

```{r, echo=TRUE}
# load mpg data set from ggplot2. mpg is a dataframe
mpg <- ggplot2::mpg

#plot displ (litres) vs. highway mpg
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy))

# plot highway mpg vs cylinders
ggplot(data = mpg) + geom_point(mapping = aes(x = hwy, y = cyl))

# plot unhelpful graph with categorical variables
ggplot(data = mpg) + geom_point(mapping = aes(class, drv))
```

### Aesthetic Mappings

```{r, echo=TRUE}
# plot with color aesthetic mapped to car class
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, color = displ<5))

# plot with size aesthetic mapped to class
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, size = year))

# plot with alpha aesthetic mapped to class
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, alpha = year))

# plot with shape aesthetic mapped to class
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, shape = class, size = class, alpha = class))

# plot and make all data points blue
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy), color = "brown1")
```

```{r, echo=TRUE}
# shows which variables are continuous and which are categorical
glimpse(mpg)
```

### Facets

```{r, echo=TRUE}
# facet wrap by class
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
    facet_wrap(~ class, nrow = 2)

# facet grid drive type and cylinders
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
    facet_grid(drv~cyl)

# facet grid cylinders
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
    facet_grid(.~cyl)

# same as above but horizontal orientation
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
    facet_grid(drv ~ .)

```

### Geometric Objects

```{r, echo=TRUE}
# left
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))

# right
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv, color = drv)) + 
    geom_point(mapping = aes(x = displ, y = hwy, color = drv))
```


```{r, echo=TRUE}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))

ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))

ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, color = drv), show.legend = FALSE)

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
    geom_smooth(mapping = aes(x = displ, y = hwy))

# cleaner
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + geom_point() + geom_smooth()

# over ride for 1 layer only
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + geom_point(mapping = aes(color = class)) + geom_smooth()

# geom smooth for subcompact only 
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
    geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)

ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_point() + geom_smooth(se = FALSE)
```

```{r, echo=TRUE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + geom_point() + geom_smooth(se = FALSE)

ggplot(data = mpg, mapping = aes(x = displ, y = hwy, class = drv)) + geom_point() + geom_smooth(se = FALSE)
  
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, col = drv)) + geom_point() + geom_smooth(se = FALSE)

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + geom_point(mapping = aes(color = drv)) + geom_smooth(se = FALSE)

ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, color = drv)) + geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv), se = FALSE)

ggplot(data = mpg, mapping = aes(x = displ, y = hwy, col = drv)) + geom_point()
```

### Statistical Transformations

```{r, echo=TRUE}
diamonds <- ggplot2::diamonds

ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))

# create using statcount instead of geom_bar
ggplot(data = diamonds) + 
  stat_count(mapping = aes(x = cut))

# override the default stat. Change stat from count to identity

demo <- tribble(
  ~a, ~b,
  "bar1", 20,
  "bar2", 30,
  "bar3", 40,
  "bar4", 50
)

ggplot(data = demo) + 
  geom_bar(mapping = aes(x = a, y = b), stat = "identity")

# display a bar chart of proportion rather than count
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = ..prop.., group = 1))

# use stat summary to give information about statistical transformations in code
ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.ymin = min,
    fun.ymax = max,
    fun.y = median
  )

#geom_pointrange
ggplot(data = diamonds) + 
  geom_pointrange(mapping = aes(x = cut, y = depth), stat = "summary")

#geom_col()
ggplot(data = diamonds) + 
  geom_col(mapping = aes(x = cut, y = depth))

``` 

### Position Adjustments

```{r, echo=FALSE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, color = cut))

ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))

# position = identity, more useful for 2d geoms
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity")

ggplot(data = diamonds, mapping = aes(x = cut, color = clarity)) + 
  geom_bar(fill = NA, position = "identity")

# position = fill, works like stacking, but makes each set of stacked bars the same height
ggplot(data = diamonds) +
  geom_bar(
    mapping = aes(x = cut, fill = clarity), position = "fill"
  )

# position = dodge places overlapping objects directly beside on another
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")

# position = jitter adds a small amount of noise to our data points to prevent overplotting and allow us to see the grouping of points
ggplot(data = mpg) + 
  geom_count(mapping = aes(x = displ, y = hwy), position = "jitter")

# default position adjustment for geom_boxplot
ggplot(data = mpg, mapping = aes(x = drv, y = hwy)) + 
  geom_boxplot()
```

### Coordinate Systems 

```{r, echo=TRUE}
#coord_flip
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + geom_boxplot()

ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + geom_boxplot() + coord_flip()
```

```{r, echo = TRUE}
#coord_quickmap
nz <- map_data("nz")

ggplot(nz, aes(long, lat, group = group)) + 
  geom_polygon(fill = "white", color = "black")

ggplot(nz, aes(long, lat, group = group)) + 
  geom_polygon(fill = "white", color = "black") + 
    coord_quickmap()
```

```{r, echo=TRUE}
#coord_polar
bar <- ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut), show.legend = FALSE, width = 1) + 
    theme(aspect.ratio = 1) + 
      labs(x = NULL, y = NULL)

bar + coord_flip()
bar + coord_polar()
```

```{r, echo=TRUE}
ggplot(data = mpg) + 
  geom_bar(mapping = aes(x = hwy, fill = drv), position = "fill") + 
    coord_polar()
```