---
title: "Untitled"
author: "Michael Rose"
date: "April 9, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Intro

You can set the output of the document:

1. Permanently by modifying the YAML header: 

```
title: "Viridis Demo"
output: html_document
```

2. Transiently, by calling rmarkdown::render() by hand:

```
rmarkdown::render(
"diamond-sizes.Rmd",
output_format = "word_document"
)
```

## Output Options

to figure out what parameters you can use, check the documentation. E.g. for html, check ?rmarkdown::html_document()

To override the deault parameter values, you need to use an extended output field. For example, if you wanted to render an html_document with a floating table of contents you'd use: 

```
output:
  html_document:
    toc: true
    toc_float: true
```

You can even render multiple outputs by supplying a list of formats: 

```
output: 
  html_document:
    toc: true
    toc_float: true
  pdf_document: default
```

## Documents

pdf_document makes a pdf typeset with latex
word_document makes a .docx
odt_document for opendocument text docs .odt
rtf_document for rich text format .rtf
md_document for a markdown document
github_document is a md_document tailored for sharing on github

When generating a document to share with decision makers, you can change global options in the setup chunk

```
knitr::opts_chunk$set(echo = FALSE)

or 

output:
  html_document:
    code_folding: hide
```

## Notebooks

You can make the ouput include both html/pdf and github documents: 

```
output:
  html_notebook: default
  github_document: default
```

## Presentations

Presentations work by dividing your content into slides with a new slide beginning at each first # or second ## header. You can also insert a horizontal rule *** to create a new slide without a header

rmarkdown comes with three presentation formats built in:

ioslides_presentation
  html presentation with ioslides
  
slidy_presentation
  html presentation with W3C slidy
  
beamer presentation
  pdf presentation with latex beamer
  
revealjs::revealjs_presentation
  html presentation with reveal.js. Requires revealjs package
  
rmdshower
  provides a wrapper around the shower presentation engine
  
## Dashboards

flexdashboard makes it easy to create dashboards

each level 1 header # begins a new page in the dashboard
each level 2 header ## begins a new column
each level 3 header ### begins a new row

for example check dmd_dist_dash.Rmd


## Interactivity

Any html format (document, notebook, presentation, or dashboard) can contain interactives.

### HTML widgets

```{r}
library(leaflet)
leaflet() %>% 
  setView(174.764, -36.877, zoom = 16) %>% 
  addTiles() %>% 
  addMarkers(174.764, -36.877, popup = "Maungawhau")

```

Packages that have widgets:

* dygraphs - interactive time series visualizations
* DT - interactive tables
* rthreejs - interactive 3d plots
* DiagrammeR - for diagrams (like flowcharts and node-link diagrams)
* htmlwidgets.org

## Shiny 

to call shiny code from an rmarkdown document:

```
title: "Shiny Web App"
output: html_document
runtime: shiny
```

ex: 

```
library(shiny)

textInput("name", "What is your name?")
numericInput("age", "How old are you?", NA, min = 0, max = 150)
```

makes a question + input box which you can map to a variable like input$age and input$name

Shiny requires a shiny server to run R. check out more at shiny.rstudio.com

## Websites 

put your .Rmd files in a single directory, index.Rmd will be the home page
add a YAML file names _site.yml to provide navigation for the site: 

```
name: "My Website"
navbar: 
  title: "My Website"
  left:
    - text: "Home"
      href: index.html
    - text: "Viridis Colors"
      href: 1-example.html
    - text: "Terrain Colors"
      href: 3-inline.html
```

execute rmarkdown::render_site() to build _site, a directory of files ready to deploy as a standalone static website.
read more at http://bit.ly/RMarkdownWebsites

## Other Formats

bookdown - makes it easy to write books. See the book Authoring Books with R Markdown orr see bookdown.org
prettydoc - lightweight document formats with a range of themes 
rticles - compiles a selection of formats for specific scientific journals

see rmarkdown.rstudio.com/formats.html for a list of more formats 



