---
title: "Chapter 5 - Exploratory Data Analysis"
author: "Michael Rose"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggstance)
library(lvplot)
library(ggbeeswarm)
library(viridis)
library(modelr)
```

```
1. What type of variation occurs within my variables?
2. What type of covariation occurs within my variables? 
```

### Visualizing Distributions

```{r, echo=TRUE}
diamonds <- ggplot2::diamonds

# categorical
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))

# manual counts
diamonds %>%
  count(cut)

# continuous
ggplot(data = diamonds) + 
  geom_histogram(mapping = aes(x = carat), binwidth = 0.5)

# manual
diamonds %>%
  count(cut_width(carat, 0.5))

# different binwidth sizes reveal different patterns
smaller <- diamonds %>% 
  filter(carat < 3)

ggplot(data = diamonds) + 
  geom_histogram(mapping = aes(x = carat), binwidth = 0.1)

# when dealing with multiple overlapping histograms, its better to use geom_freqpoly
ggplot(data = diamonds, mapping = aes(x = carat, color = cut)) + 
  geom_freqpoly(binwidth = 0.1)
```

### Typical Values

Heights of bars and lines show the abundance or lack thereof of particular values. To turn this into useful questions, consider:

* Which values are the most common and why?
* Which values are rare? Why? Does that match your expectations?
* Can you see any unusual patterns? What might explain them? 

Examples for the Diamonds data set:

* Why are there more diamonds at whole carats and common fractions of carats? 
* Why are there more diamonds slightly to the right of each peak than there are slightly to the left of each peak? 
* Why are no diamonds bigger than 3 carats? 

```{r, echo = TRUE}
ggplot(data = smaller, mapping = aes(x = carat)) + 
  geom_histogram(binwidth = 0.01)
```

In general, clusters of similar values suggest that subgroups exist in the data. To understand the subgroups, ask:

* How are the observations within each cluster similar to each other? 
* How are the observations in seperate clusters different from each other? 
* How can you explain or describe the clusters? 
* Why might the appearance of clusters be misleading? 


The following histogram shows the length (in minutes) of 272 eruptions of the old faithful geyser in Yellowstone National Park. There are 2 clusters, or groupings of eruptions: those around 2 minutes and those around 4-5 minutes. 
```{r}
ggplot(data = faithful, mapping = aes(x = eruptions)) + 
  geom_histogram(binwidth = 0.25)
```

### Unusual Values

```{r, echo=TRUE}

# notice the long x axis. This is indicative of an outlier. 
ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = y), binwidth = 0.5)

# make it easier to see
ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = y), binwidth = 0.5) + 
  coord_cartesian(ylim = c(0, 50))

# pluck out the data points with dplyr
unusual <- diamonds %>%
  filter(y < 3 | y > 20) %>%
  arrange(y)
unusual

# ctrl alt space %>% 

# explore the distribution of x, y, and z

diamonds %>% 
  mutate(id = row_number()) %>% 
  select(x, y, z, id) %>% 
  gather(variable, value, -id) %>% 
  ggplot(aes(x = value)) +
  geom_density() + 
  geom_rug() + 
  facet_grid(variable ~.)

# explore the distribution of price

ggplot(diamonds) +
  geom_histogram(mapping = aes(x = price), binwidth = 1000)

ggplot(filter(diamonds, price < 2500), aes(x = price)) + 
  geom_histogram(binwidth = 10, center = 0)

ggplot(filter(diamonds), aes(x = price)) + 
  geom_histogram(binwidth = 100, center = 0)

# Distributionn of last digit
diamonds %>% 
  mutate(ending = price %% 10) %>% 
  ggplot(aes(x = ending)) + 
  geom_histogram(binwidth = 1, center = 0) + 
  geom_bar()

diamonds %>% 
  mutate(ending = price %% 100) %>% 
  ggplot(aes(x = ending)) + 
  geom_histogram(binwidth = 1) + 
  geom_bar()

diamonds %>% 
  mutate(ending = price %% 1000) %>% 
  filter(ending >= 500, ending <= 800) %>% 
  ggplot(aes(x = ending)) + 
  geom_histogram(binwidth = 1) + 
  geom_bar()

# How many diamonds are 0.99 carat? How many are 1 carat?

diamonds %>% 
  filter(carat >= 0.99, carat <= 1) %>% 
  ggplot(aes(x = carat)) + 
  geom_bar()

#coord_cartesian vs. xlim, ylim

ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = price)) + 
  coord_cartesian(xlim = c(100, 5000), ylim = c(0, 3000))

# coord_cartesian zooms in on the area specified by the limits. The calculation of the histogram is unaffected. 

# xlim and ylim functions drop any values outside the limits, then calculate the histogram and then draws the graph within the given limits.  

ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = price)) +
  xlim(100, 5000) +
  ylim(0, 3000)
```

### Missing Values

```{r}
# if we encounter unusual values, it is better to make them into NA values than to delete them 

# drop the entire row (don't do):
# diamonds2 <- diamonds %>% filter(between(y, 3, 20))

# replace with NA (recc'd):
diamonds2 <- diamonds %>% 
  mutate(y = ifelse(y <3 | y > 20, NA, y))

ggplot(data = diamonds2, mapping = aes(x = x, y = y)) + 
  geom_point(na.rm = TRUE)

# to compare values that are na with values that are not na

nycflights13::flights %>% 
  mutate(
    cancelled = is.na(dep_time),
    sched_hour = sched_dep_time %/% 100, 
    sched_min = sched_dep_time %% 100, 
    sched_dep_time = sched_hour + sched_min / 60
  ) %>% 
  ggplot(mapping = aes(sched_dep_time)) + 
    geom_freqpoly(
      mapping = aes(color = cancelled),
      binwidth = 1/4
    )

# Missing values are removed in a histogram. 

ggplot(diamonds2, aes(x = y)) + 
  geom_histogram()

# in geom_bar, NA is considered a category

diamonds %>% 
  mutate(cut = if_else(runif(n()) < 0.1, NA_character_, as.character(cut))) %>% 
  ggplot() + 
  geom_bar(mapping = aes(x = cut))
```

### Covariation

Covariance describes the behavior between variables. Covariation is the tendency for the values or 2 or more variables to vary together in a related way. 

```{r}
# check how the price of a diamond varies with its quality
ggplot(data = diamonds, mapping = aes(x = price)) + 
  geom_freqpoly(mapping = aes(color = cut), binwidth = 500)

# see difference in counts
ggplot(diamonds) + geom_bar(mapping = aes(x = cut))

# swap axis. density is the count standardized so that the area under each frequency poly is 1. 

ggplot(data = diamonds, mapping = aes(x = price, y = ..density..)) + 
  geom_freqpoly(mapping = aes(color = cut), binwidth = 500)

# box plot

ggplot(data = diamonds, mapping = aes(x = cut, y = price)) + 
  geom_boxplot()

# cut is an ordered factor, but when the factors aren't ordered we can use reorder()

# original 
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + geom_boxplot()

# reordered

ggplot(data = mpg) + 
  geom_boxplot(mapping = aes(
    x = reorder(class, hwy, FUN = median),
    y = hwy
  ))

# with long var names, flip the boxplot

ggplot(data = mpg) + 
  geom_boxplot(
    mapping = aes(
      x = reorder(class, hwy, FUN = median),
      y = hwy
    )
  ) + 
  coord_flip()

# improving departure times of cancelled and non-cancelled flights

nycflights13::flights %>% 
  mutate(
    cancelled = is.na(dep_time),
    sched_hour = sched_dep_time %/% 100,
    sched_min = sched_dep_time %% 100,
    sched_dep_time = sched_hour + sched_min / 60
  ) %>% 
  ggplot() + geom_boxplot(mapping = aes(x = cancelled, y = sched_dep_time))

# What variables in the diamonds dataset is most important for predicting the price of a diamond? How is that variable correlated with cut? Why does the combination of those two relationships lead to lower quality diamonds being more expensive? 

# sounds like some regression is needed

# horizontal boxplot

# regular 

ggplot(data = mpg) + 
  geom_boxplot(mapping = aes(x = reorder(class, hwy, FUN = median), y = hwy)) + coord_flip()

# ggstance

ggplot(data = mpg) + 
  geom_boxploth(mapping = aes(y = reorder(class, hwy, FUN = median), x = hwy))

# lvplots show more quantiles and look quite nice. The more quantiles part makes them more useful for larger data sets. 

ggplot(data = diamonds) + 
  geom_lv(mapping = aes(y = price, x = cut))

# faceted geom_histogram

ggplot(data = diamonds, mapping = aes(x = price)) + 
  geom_histogram() + facet_wrap(~ cut, ncol = 1, scales = "free_y")

# violin

ggplot(data = diamonds, mapping = aes(x = cut, y = price)) + 
  geom_violin() + coord_flip()

# ggbeeswarm

# geom_quasirandom

ggplot(data = mpg) + 
  geom_quasirandom(mapping = aes(x = reorder(class, hwy, FUN = median), y = hwy))

ggplot(data = mpg) + 
  geom_quasirandom(mapping = aes(x = reorder(class, hwy, FUN = median), 
                                 y = hwy),
                   method = "tukey")

ggplot(data = mpg) + 
  geom_quasirandom(mapping = aes(x = reorder(class, hwy, FUN = median), y = hwy),
                   method = "tukeyDense")

ggplot(data = mpg) + 
  geom_quasirandom(mapping = aes(x = reorder(class, hwy, FUN = median), y = hwy), method = "frowney")

ggplot(data = mpg) + 
  geom_quasirandom(mapping = aes(x = reorder(class, hwy, FUN = median), y = hwy), method = "smiley")

# geom_beeswarm

ggplot(data = mpg) + 
  geom_beeswarm(mapping = aes(x = reorder(class, hwy, fun = median), y = hwy))
```
 
### Two Categorical Variables

```{r}
# count the number of observations for each combination of cut and color

ggplot(data = diamonds) + 
  geom_count(mapping = aes(x = cut, y = color))

# count with dplyr

diamonds %>% 
  count(color, cut) %>% 
  ggplot(mapping = aes(x = color, y = cut)) + 
  geom_tile(mapping = aes(fill = n))

# How could you rescale the count dataset to more clearly show the distribution of cut within color, or color within cut? 

# calc a new var, prop which shows proportion of each cut within a color

diamonds %>% 
  count(color, cut) %>% # make counts of colors
  group_by(color) %>% # create color groups
  mutate(prop = n / sum(n)) %>%  # create prop var
  ggplot(mapping = aes(x = color, y = cut)) + # ggplot
  geom_tile(mapping = aes(fill = prop)) + 
  scale_fill_viridis() # color

# scale by color within a cut

diamonds %>% 
  count(color, cut) %>% # make counts of colors
  group_by(cut) %>% # create color groups
  mutate(prop = n / sum(n)) %>%  # create prop var
  ggplot(mapping = aes(x = color, y = cut)) + # ggplot
  geom_tile(mapping = aes(fill = prop)) + 
  scale_fill_viridis() # color


# How do average flight delays vary by destination and month of year

flights <- nycflights13::flights

# looks like shit

flights %>% 
  group_by(month, dest) %>% 
  summarize(dep_delay = mean(dep_delay, na.rm = TRUE)) %>% 
  ggplot(aes(x = factor(month), y = dest, fill = dep_delay)) + 
  geom_tile() + 
  labs(x = "Month", y = "Destination", fill = "Departure Delay")

# less so

flights %>%
  group_by(month, dest) %>% 
  summarize(dep_delay = mean(dep_delay, na.rm = TRUE)) %>% 
  group_by(dest) %>% 
  filter(n() == 12) %>% 
  ungroup() %>% 
  mutate(dest = fct_reorder(dest, dep_delay)) %>% 
  ggplot(aes(x = factor(month), y = dest, fill = dep_delay)) + 
  geom_tile() + 
  scale_fill_viridis() + 
  labs(x = "Month", y = "Destination", fill = "Departure Delay")
  

# Why is it better to use aes(x = color, y = cut) rather than aes(x = cut, y = color)

# Interpretability

diamonds %>%
  count(color, cut) %>%
  ggplot(mapping = aes(y = color, x = cut)) + 
    geom_tile(mapping = aes(fill = n))
  
```

### Two Continuous Variables

```{r}
# relationship between carat size and price

ggplot(data = diamonds) + 
  geom_point(mapping = aes(x = carat, y = price))

# to deal with the blackout effect of large datasets, we can use alpha

ggplot(data = diamonds) + 
  geom_point(mapping = aes(x = carat, y = price), 
             alpha = 1/100)
```

#### bins

```{r}
#geom_bin2d

ggplot(data = smaller) + 
  geom_bin2d(mapping = aes(x = carat, y = price))

# geom_hex

ggplot(data = smaller) + 
  geom_hex(mapping = aes(x = carat, y = price))

# bin one continuous variable to make it act like a categorical variable

ggplot(data = smaller, mapping = aes(x = carat, y = price)) + 
  geom_boxplot(mapping = aes(group = cut_width(carat, 0.1)))

# boxplots don't show the proportion of points within well. We can display approximately the same number of points within each bin like so: 

ggplot(data = smaller, mapping = aes(x = carat, y = price)) + 
  geom_boxplot(mapping = aes(group = cut_number(carat, 20)))

# Instead of summarizing the conditional distribution with a boxplot, you could use a frequency polygon. What do you need to consider when using cut_width(), vs cut_number()? How does that impact a visualization of the 2d distribution of carat and price? 

# with cut_width, the number of points in the bin may not be equal

ggplot(data = diamonds, mapping = aes(x = price, color = cut_width(carat, 0.3))) + geom_freqpoly()

# plotting the density instead of counts makes distributions comparable, but bins with few observations will still be hard to interpret

ggplot(data = diamonds, mapping = aes(x = price, y = ..density..,
                                      color = cut_width(carat, 0.3))) + 
  geom_freqpoly()

# adjusting the bin size

ggplot(data = diamonds, 
       mapping = aes(x = price,
                     colour = cut_number(carat, 10))) +
  geom_freqpoly()

# with density

ggplot(data = diamonds, 
       mapping = aes(x = price, y = ..density..,
                     color = cut_number(carat, 10))) + 
  geom_freqpoly()

# visualize the distribution of carat, partitioned by price

ggplot(data = diamonds,
       mapping = aes(x = carat, y = price)) + 
  geom_boxplot(mapping = aes(group = cut_width(price, 1000)))

ggplot(data = diamonds, aes(x = cut_number(price, 10), y = carat)) + 
  geom_boxplot() + coord_flip() + xlab("price")

# Combine two of the techniques you've learned to visualize the combined distribution of cut, carat, and price

# 1

ggplot(data = diamonds, aes(x = carat, y = price)) + 
  geom_hex() + 
  facet_wrap(~ cut, ncol = 1) + 
  scale_fill_viridis()

# 2

ggplot(data = diamonds, aes(x = cut_number(carat, 5), y = price, color = cut)) + 
  geom_boxplot()

# 3

ggplot(data = diamonds, mapping = aes(x = cut_number(carat, 1), y = price, color = cut)) + geom_violin()

# 4

ggplot(data = diamonds, mapping = aes(x = carat, y = price, color = cut)) + geom_point()

```

### Patterns and Models

```

When spotting a pattern, ask:

* Could this be due to coincedence? 
* How can you describe the relationship?
* How strong is the relationship?
* What other variables might affect the relationship?
* Does the relationship change if you look at individual subgroups of the data? 

```

```{r}

# plot old faithful. Notice the grouping. Clustering likely works best
ggplot(data = faithful) + 
  geom_point(mapping = aes(x = eruptions, y = waiting))

# make log price vs log carat model
mod <- lm(log(price) ~ log(carat), data = diamonds)

# add residuals to data set
diamonds2 <- diamonds %>% 
  add_residuals(mod) %>% 
  mutate(resid = exp(resid))

# plot residuals

ggplot(data = diamonds2) + geom_point(mapping = aes(x = carat, y = resid))

# plot boxplot
ggplot(data = diamonds2) + 
  geom_boxplot(mapping = aes(x = cut, y = resid))
```

### ggplot2 calls

```{r}

# current
ggplot(data = faithful, mapping = aes(x = eruptions)) + 
  geom_freqpoly(binwidth = 0.25)

# more concisely

ggplot(faithful, aes(eruptions)) + 
  geom_freqpoly(binwidth = 0.25)
```

